Enter the \texttt{holodeck}: massive black hole binary population synthesis for gravitational wave calculations

### 2023-06-19 docs/holodeck-paper/ms.tex

\newtotcounter{citnum} % From the package documentation
\def\oldbibitem{} \let\oldbibitem=\bibitem
\def\bibitem{\stepcounter{citnum}\oldbibitem}

\newcommand{\orcidauthorA}{0000-0002-6625-6450} % For author A

@@

    Massive black holes (MBHs) occupy the centers of most if not all massive galaxies \needcite{}.
    Binaries of MBHs are believed to form following the merger of massive galaxies \needcite{}.  Such MBH Binaries (MBHBs) could then produce low-frequency gravitational waves (GWs) \needcite{LF-GWs}, in the nano-Hertz regime, that would soon be detectable by experiments called pulsar timing arrays (PTAs) \needcite{PTAs}.  In this paper, we introduce the MBHB population synthesis code called \holodeck{}.  The goal of this package is to provide a cohesive and comprehensive framework for:
    \begin{enumerate}
        \item
        modeling the physics of MBHB formation and evolution while accurately capturing the true uncertainties in this modeling;
        \item
        rapidly generating large simulated populations of MBHBs;
        \item
        calculating the resulting GW signals from these populations.
    \end{enumerate}
    % (1) modeling the physics of MBHB formation and evolution while accurately capturing the true uncertainties in this modeling, (2) rapidly generating large simulated populations of MBHBs, and (3) calculating the resulting GW signals from these populations.

@@

        The number-density of binaries is conserved, which allows us to take finite steps in separation and time/redshift without approximation: $a\rightarrow a'$ and $z\rightarrow z'$.  The time it takes for a binary to go between separations $a_1 \rightarrow a_2$ is
        % \mbox{$T \equiv \int_{a_1}^{a_2} \frac{da}{\partial a / \partial t}$}.
        \mbox{$T \equiv \int_{a_1}^{a_2} da / \lr{\partial a / \partial t}$}.
        This leads to a redshift at the later time of $z' = z'(t + T)$.  This allows us to approximate the term from \eqref{eq:conservation_fixed_mass} as,
        \begin{equation}
            \label{eq:}
            \diffp{}{a} \lrs{F(M,q,z,a) \diffp{a}{t}} \approx \frac{F(M,q,z,a)}{T_g(m_1,q,z)}.
        \end{equation}
        The approximation is tied to the marginalization over binary separation in the GMT term.  Because the number density of binaries $n$ is not explicitly dependent on the separation, we can still write exactly that
        \begin{equation}
            \label{eq:sam_gpf_fmt}
            \diffp{n}{{M}{q}{z'}} = - \diffp{t}{z} \frac{\Phi_g(m_1,z) \, P_g(m_1,q,z)}{T_g(m_1,q,z)}.
        \end{equation}
        Putting this all together, we can write our SAM number of binaries as either,
        \begin{equation}
            \begin{split}
                \label{eq:sam_final_gmr}
                \diffp{N}{{M}{q}{z'}{\ln \! f_r}} & = \Phi_g(m_1,z) \, R_g(m_1,q,z) \, \thardf(M,q,z',f) \, \diffp{{V_c}}{{z'}}, \\
                & = \Phi_g(m_1,z) \, \frac{P_g(m_1,q,z)}{T_g(m_1,q,z)} \, \thardf(M,q,z',f) \, \diffp{V_c}{{z'}}.
            \end{split}
        \end{equation}
        \todo{Should $R_g$ use $z$ or $z'$ (redshift-prime)?}

>>>>

% \newtotcounter{citnum} % From the package documentation
% \def\oldbibitem{} \let\oldbibitem=\bibitem
% \def\bibitem{\stepcounter{citnum}\oldbibitem}

% \newcommand{\orcidauthorA}{0000-0002-6625-6450} % For author A

@@

    Massive black holes (MBHs) occupy the centers of most if not all massive galaxies \needcite{}.
    Binaries of MBHs are believed to form following the merger of massive galaxies \needcite{}.  Such MBH Binaries (MBHBs) could then produce low-frequency gravitational waves (GWs) \needcite{LF-GWs}, in the nano-Hertz regime, that would soon be detectable by experiments called pulsar timing arrays (PTAs) \needcite{PTAs}.  In this paper, we introduce the MBHB population synthesis code called \holodeck{}.  The goal of this package is to provide a cohesive and comprehensive framework for:
    % Massive black holes (MBHs) are known to occupy the centers of most if not all massive galaxies \needcite{}.
    Binaries of MBHs form following the merger of massive galaxies \needcite{}.  These MBH Binaries (MBHBs) can then produce the loudest gravitational waves (GWs) in the Universe.  MBHBs with total masses $M \gtrsim 10^8 \,\msol$ are detectable by experiments called pulsar timing arrays (PTAs) \needcite{PTAs} in the nanohertz regime \needcite{}, while more modest systems with $M \lesssim 10^7 \, \msol$ will be detectable by LISA \needcite{LISA} at millihertz \needcite{}.  While accreting, MBHs are visible across the electromagnetic (EM) spectrum as active galactic nuclei (AGN) \needcite{}, suggesting that MBHBs could be promising multimessenger sources \needcite{}.  A large number of studies have modeled cosmological populations of MBHBs, and made predictions for their GW signatures and EM counterparts, however no comprehensive framework has been released.  In this paper, we introduce the MBHB population synthesis code \holodeck{}, developed within the astrophysics working group of the NANOGrav\footnote{North American nanohertz observatory for gravitational waves \citep{}: \url{https://www.nanograv.org}.}.  The goal of this package is to provide a cohesive and comprehensive framework for:
    \begin{enumerate}
        \item
        modeling the physics of MBHB formation and evolution, with parameterizations for uncertain physical processes;
        \item
        rapidly generating large simulated populations of MBHBs; and,
        \item
        calculating the resulting GW signals and EM signatures from these populations.
    \end{enumerate}
    % (1) modeling the physics of MBHB formation and evolution while accurately capturing the true uncertainties in this modeling, (2) rapidly generating large simulated populations of MBHBs, and (3) calculating the resulting GW signals from these populations.

@@

        The number-density of binaries is conserved, which allows us to take finite steps in separation and time/redshift without approximation: $a\rightarrow a'$ and $z\rightarrow z'$.  The time it takes for a binary to go between separations $a_1 \rightarrow a_2$ is
        % \mbox{$T \equiv \int_{a_1}^{a_2} \frac{da}{\partial a / \partial t}$}.
        \mbox{$T \equiv \int_{a_1}^{a_2} da / \lr{\partial a / \partial t}$}.
        This leads to a redshift at the later time of $z' = z'(t + T)$.  This allows us to approximate the term from \eqref{eq:conservation_fixed_mass} as,
        \begin{equation}
            \label{eq:}
            \diffp{}{a} \lrs{F(M,q,z,a) \diffp{a}{t}} \approx \frac{F(M,q,z,a)}{T_g(m_1,q,z)}.
        \end{equation}`'
        The approximation is tied to the marginalization over binary separation in the GMT term.  Because the number density of binaries $n$ is not explicitly dependent on the separation, we can still write exactly that
        \begin{equation}
            \label{eq:sam_gpf_fmt}
            \diffp{n}{{M}{q}{z'}} = - \diffp{t}{z} \frac{\Phi_g(m_1,z) \, P_g(m_1,q,z)}{T_g(m_1,q,z)}.
        \end{equation}
        Putting this all together, we can write our SAM number of binaries as either,
        \begin{equation}
            \begin{split}
                \label{eq:sam_final_gmr}
                \diffp{N}{{M}{q}{z'}{\ln \! f_r}} & = \Phi_g(m_1,z) \, R_g(m_1,q,z) \, \thardf(M,q,z',f) \, \diffp{{V_c}}{{z'}}, \\
                & = \Phi_g(m_1,z) \, \frac{P_g(m_1,q,z)}{T_g(m_1,q,z)} \, \thardf(M,q,z',f) \, \diffp{V_c}{{z'}}.
            \end{split}
        \end{equation}
        \todo{Should $R_g$ use $z$ or $z'$ (redshift-prime)?}
